Inter-rater reliability of the tracking and reporting outcomes of procedural sedation tool

### 2021-06-17 response.Rmd

---
output: 
    pagedown::html_paged:
        css: styles.css
knit: pagedown::chrome_print
---

Thank you for allowing us the opportunity to revise our manuscript. Below are the reviewers' comments and our responses.

**Reviewer 1**

*General Overview*

> Thank you for the opportunity to review "Inter-rater reliability of the tracking and reporting outcomes of procedural sedation tool," submitted for consideration to the Journal of Radiology Nursing. This manuscript addresses a very interesting area of research and discusses an important topic with regard to outcomes for procedural sedation.

*Other Comments*

>1.The manuscript is well organized overall with regard to the clear sections denoting the introduction, stated aims, methods, and results as examples.

Thank you.

>2.Introduction: It may strengthen the introduction of the paper if details related to the impact of sedation-related complications were added.

Added to the introduction:

'Serious adverse effects of sedation, such as respiratory compromise, are associated with worse clinical outcomes and higher costs.'

>3.Line 53, Participants: It is recommended by this reviewer that 'nurse - administered sedation' is specifically defined for the reader as this concept can have multiple meanings.

Added the following:

'Nurse-administered sedation' is when nurses administer a combination of sedative and analgesic medications prescribed by the proceduralist performing the procedure.

>4.Line 16, Measurement: There is a brief description of the TROOPS tool provided in the introduction however, it is unclear as to whether the TROOPS tool is a validated measure for groups other than nursing. If it is a validated measure there is concern related to adding and subtracting questions and how this impacts the overall validity of the measure.

Added:

The TROOPS tool has not undergone validity or reliability assessment in any other clinical setting.

>5.Line 45, Measurement: It appears that some modifications were made to the TROOPS tool for purposes of this study testing the feasibility and reliability for use by nurses. By making changes to the TROOPS QI tool, the feasibility and reliability of the TROOPS QI tool was not necessarily tested. Instead this reviewer would suggest the authors tested the feasibility and reliability of a modified version of the tool they created for the study, not the actual TROOPS QI tool itself. Additional studies assessing feasibility and reliability for nursing would need to use the modified version creating by these authors.

We agree with this statement. In the original manuscript we used the phrase 'version of the TROOPS tool' to make the distinction that the exact tool was not used. In this revision, we instead add the term 'modified' to make the distinction clearer. We made this change in the manuscript and throughout the main body of the manuscript.

>6.Line 19, Data collection: If possible please define the role of 'scout' and 'monitor.'

Added:

The nurse performing the 'scout' role was responsible for administering direct patient care and procedural support. The nurse performing the 'monitor' role was responsible for physiological monitoring and related documentation.

>7.Line 19, Is it possible that ratings between nurses using the TROOPS tool were not in agreement due to the modifications? Has IRR been calculated using the tool in other scenarios?

The IRR of the tool has not been evaluated previously. But we agree that our estimates of IRR are only relevant to the modified version of the tool that we used in the study. We added this sentence to the limitations section:

'Finally, estimates of inter-rater reliability reported in this study should only be considered applicable to the modified version of the tool that we used, not the original research or QI versions of the TROOPS tool.'

**Reviewer #2**

>Summary - The researchers conducted a quality improvement check on the interrater reliability of the TROOPS tool to be used in the cardiac catherization lab for procedural sedation outcomes. Overall, the manuscript is clearly written and does not overreach conclusions of results. There are some points of clarity that are outlined below.

Thank you

:::{.break}
>Abstract - The abstract was comprehensive. It may have been helpful for the abstract to be structured following the research headings of background, purpose, methods, results, conclusion. The reader was unclear of the purpose of the manuscript until the last sentence of the abstract.
:::

We decided not to change to a structured style for the abstract, as the unstructured style seems to be used for other research articles published in the Journal of Radiology Nursing.

>The last sentence of the abstract - using the word "seems" does not provide confidence to the reader that the TROOPS tool is one that should be adopted. Suggest a rewrite.

Changed sentence to:

"Use of this tool in clinical practice is feasible and reliable."

> Purpose and Aims: The purpose and aims statements are clear and easy to understand. For improved readability suggest that aim 'b' be revised to read……..calculate to determine the impact that integrating auditing of sedation outcomes into routine clinical practice on nurse workload).

Bullet point changed to:

Time taken for nurses to complete the TROOPS tool (calculated to determine the impact that integrating auditing of sedation outcomes into routine clinical practice on nurse workload).

>  Participants: Suggest providing your n=40 patients here in this section. This will make it clearer for the reader to understand the data collection and data analysis sections.

We believe that the number of participants included in the study is better suited to be presented in the results section. No changes were made.

>Data collection: The terms scout and monitor need to be defined for the readership. These are not universally understood terms or roles internationally, nor outside cardiac catheterization.

Added:

The nurse performing the 'scout' role was responsible for administering direct patient care and procedural support. The nurse performing the 'monitor' role was responsible for physiological monitoring and documentation.

>Discussion: Given that the results showed minor adverse effects in 50% of your patients, I wonder if the point should be made in the discussion that the anticipation of potential minor adverse effects such as the need for oxygen support or antiemetics during procedural sedations is the very reason these patients are 1:1 nursing surveillance during their sedation. The finding of a minor adverse effect such as was reported does not surprise me. Missing however in the discussion is what was done about these findings at the clinical level during the sedations. Minor adverse events escalate into intermediate or serious adverse events if not recognized and treated while still minor. Therefore, somewhat concerning is that only one rater reported the adverse event and the other missed that it occurred. I think it would be more interesting to the readership to have some discussion on this point. Consistency of nursing surveillance improves patient outcomes.

We do of course agree that nursing surveillance is vital. However, the design of this study was solely focused on investigating the inter-rater reliability and feasibility of using the TROOPS tool for nurse-administered sedation in the cath lab. We would prefer to avoid discussing possible clinical implications of the ~50% minor adverse event rate, because we have no way of knowing that this is an accurate estimate of the true event rate. As stated in the limitations section, estimating the prevalence of adverse sedation-related events was not the primary purpose of this study so we did not collect information about the total number of procedures performed with sedation over the data collection period to determine the exact proportion of completed TROOPS ratings.

### 2021-06-17 manuscript.Rmd
R1

Clinical guidelines recommend that procedural sedation outcomes should be audited. The Tracking and Reporting Outcomes of Procedural Sedation (TROOPS) tool, developed by the International Consortium for the Advancement of Procedural Sedation, is intended to be suitable for use in all locations by sedation providers. We evaluated the feasibility and reliability of using the TROOPS tool for sedation that is administered by nurses in the cardiac catheterization laboratory. The version of TROOPS we used contained items in the minor, intermediate and sentinel categories. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A sedation-related adverse event was selected by at least one of the raters for 21 of the patients (52%; 95% CI = 0.38 to 0.67). Most were minor events related to the airway and breathing category of the TROOPS tool (n = 17; 42%; 95% CI = 0.29 to 0.58). The remaining were intermediate-severity events related to sedation quality. No events for the other categories of the TROOPS tool were selected. The intra-class correlation coefficient between paired nurse ratings of the TROOPS tool was `r total` (95% CI = `r total_lower` to `r total_upper`). Most nurses (85%, 95% CI = 73% to 92%) reported that it took less than 1 minute to complete the tool. In summary, paired ratings of sedation-related adverse events by nurses achieved moderate consistency using a version of the TROOPS tool containing items from the minor category of the airway and breathing domain. Use of this tool in clinical practice seems feasible.

@@

The administration of sedative and analgesic medications during medical procedures is associated with several unintended effects with the potential to compromise patient safety. However, if systems are in place to manage these complications, serious sedation-related serious adverse events are preventable. One central aspect of ensuring patient safety during sedation that is recommended in clinical guidelines is auditing of sedation-related outcomes.[@Dobson2018; @apfelbaum2018; @werthman2020administration; @hinkelbein2018; @conway2013] For example, the Practice Guidelines for Moderate Procedural Sedation and Analgesia from the American Society of Anesthesiologists Task Force in Moderate Procedural Sedation recommend the creation and implementation of a quality improvement process to measure adverse events and unsatisfactory sedation.[@apfelbaum2018] Likewise, the Canadian Anesthesiologists Society recommends that a specific mechanism must be established for auditing procedural sedation.[@Dobson2018]

@@

1.	Inter-rater reliability of nurses’ TROOPS ratings; and
2.	Time taken for nurses to complete the TROOPS tool (calculated to determine the impact that integrating auditing of sedation outcomes into routine clinical practice would have on nurses’ workload).

@@

Patients undergoing procedures with nurse-administered sedation in a cardiac catheterization laboratory were eligible. A convenience sample was used.

@@

The second part consisted of the quality improvement version of the TROOPS tool, with the addition of minor adverse events from the 'airway and breathing' as well as 'gastrointestinal' domains from the research version of the tool.[@roback2018tracking] Minor events from the 'airway and breathing' domain used were 'increased or added supplemental oxygen', 'airway repositioning' and 'tactile stimulation'. Based on consultation with members of the clinical team, we elected not to include the minor items related to hypersalivation (i.e., suctioning for hypersalivation and anticholinergic for hypersalivation) because the medications used for sedation in the department (i.e., fentanyl and midazolam) are not associated with this particular adverse effect. Both minor events from the gastrointestinal domain of the research version were used. These were administration of anti-emetic and suction for emesis. The minor items from the allergy and neurological domains were not included because these adverse effects were not considered to occur frequently enough in this particular sedation context to warrant inclusion as a measure for quality improvement. Other modifications were made to the TROOPS tool through consultation with members of the clinical team in order to make it suitable for use by nurses in that particular department. Modifications included the removal of outcomes that were not feasibly assessed by nurses in the immediate post-procedure period, including permanent neurological deficit from the sentinel category of the neurological domain and pulmonary aspiration syndrome from the sentinel adverse events category of the airway and breathing domain.

@@

Two nurses who performed the ‘scout’ and ‘monitor’ role during procedures performed with nurse-administered sedation completed the TROOPS tool. Ratings were performed independently as soon as possible after the completion of the procedure. Prior to commencement of the project, all nurses involved in administering sedation in the clinical department were invited to attend in-services where information about the project and the assessment tool was provided. We did not collect information on the number of nurses attending the education sessions.

@@

Data were collected from January to November, 2017. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A summary of the patient characteristics is presented in Table \@ref(tab:tab1). Most patients were undergoing a cardiac implantable electronic device implant (e.g. pacemaker or implantable cardioverter defibrillator) or electrophysiology procedure (e.g., atrial fibrillation or ventricular tachycardia ablation). A combination of midazolam and fentanyl was used for all patients.

A sedation-related adverse event was selected by at least one of the raters for 21 of the patients (52%; 95% CI = 0.38 to 0.67). Most were minor events related to the airway and breathing category of the TROOPS tool (n = 17; 42%; 95% CI = 0.29 to 0.58). The remaining were intermediate-severity events related to sedation quality and patient experience. No events for the other categories of the TROOPS tool were selected. The intra-class correlation coefficient between paired nurse ratings of the TROOPS tool was `r total` (95% CI = `r total_lower` to `r total_upper`). Responses from fifty-one (85%, 95% CI = 73% to 92%) nurses indicated that it took less than 1 minute to complete the TROOPS tool. The remaining nurses responded that it took between 1 and 3 minutes to complete. There were 20 missing responses to this item for the 40 patients.

@@

Reliability is a vital property of a measurement tool if results are intended to inform quality improvement.  There was moderate consistency between paired TROOPS ratings. However, it should be noted that confidence intervals for the intra-class correlation were wide, spanning from poor (`r total_lower`) to very high levels of consistency (`r total_upper`). Additional research with a larger sample size would increase confidence in estimates of inter-rater reliability for the TROOPS tool.

@@

We did not collect information on which nurses completed TROOPS ratings so that anonymity was maintained. As such, the total number of nurses who completed TROOPS ratings for patients in this study is not known. For the same reason, we can not report on the characteristics of the nurses who provided ratings, such as their experience or training.[@kottner2011guidelines] It should also be noted that, as expected, most events identified were from the minor category of the airway and breathing domain. Results should not be generalized outside of this context. The less common and more severe adverse events, such as the unintended requirement for tracheal intubation during sedation, is not likely to be inconsistently rated by nurses. Likewise, results should be considered most applicable to the context in which sedation was performed in this study. An approach that is common across many cardiac catheterization laboratories was used, with bolus doses of midazolam and fentanyl being administered by nurses.[@Conway_2014] Although there was a substantial proportion of events recorded by nurses in the sample (52%), it is possible that it over-estimates the true event rate because the analysis was performed using a convenience sample. The sample represents only a small proportion of the total number of procedures performed with sedation at the site over the data collection period. Estimating the prevalence of adverse sedation-related events was not the primary purpose of this study so we did not collect information about the total number of procedures performed with sedation over the data collection period to determine the exact proportion of completed TROOPS ratings.

@@

Paired ratings of sedation-related adverse events by nurses in the cardiac catheterization laboratory achieved moderate consistency using a version of the TROOPS tool containing items from the minor category of the airway and breathing domain. Use of this tool in clinical practice seems feasible. The time required to complete the ratings was very minimal, with most ratings taking less than one minute to complete.

>>>>

Clinical guidelines recommend that procedural sedation outcomes should be audited. The Tracking and Reporting Outcomes of Procedural Sedation (TROOPS) tool, developed by the International Consortium for the Advancement of Procedural Sedation, is intended to be suitable for use in all locations by sedation providers. We evaluated the feasibility and reliability of using a modified version of the TROOPS tool for sedation that is administered by nurses in the cardiac catheterization laboratory. The version of TROOPS we used contained items in the minor, intermediate and sentinel categories. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A sedation-related adverse event was selected by at least one of the raters for 21 of the patients (52%; 95% CI = 0.38 to 0.67). Most were minor events related to the airway and breathing category of the TROOPS tool (n = 17; 42%; 95% CI = 0.29 to 0.58). The remaining were intermediate-severity events related to sedation quality. No events for the other categories of the TROOPS tool were selected. The intra-class correlation coefficient between paired nurse ratings of the TROOPS tool was `r total` (95% CI = `r total_lower` to `r total_upper`). Most nurses (85%, 95% CI = 73% to 92%) reported that it took less than 1 minute to complete the tool. In summary, paired ratings of sedation-related adverse events by nurses achieved moderate consistency using a modified version of the TROOPS tool containing items from the minor category of the airway and breathing domain. Use of this tool in clinical practice is feasible and reliable.

@@

The administration of sedative and analgesic medications during medical procedures is associated with several unintended effects with the potential to compromise patient safety. Serious adverse effects of sedation, such as respiratory compromise, are associated with worse clinical outcomes and higher costs.[@urman2019impact] However, if systems are in place to manage these complications, serious sedation-related serious adverse events are preventable. One central aspect of ensuring patient safety during sedation that is recommended in clinical guidelines is auditing of sedation-related outcomes.[@Dobson2018; @apfelbaum2018; @werthman2020administration; @hinkelbein2018; @conway2013] For example, the Practice Guidelines for Moderate Procedural Sedation and Analgesia from the American Society of Anesthesiologists Task Force in Moderate Procedural Sedation recommend the creation and implementation of a quality improvement process to measure adverse events and unsatisfactory sedation.[@apfelbaum2018] Likewise, the Canadian Anesthesiologists Society recommends that a specific mechanism must be established for auditing procedural sedation.[@Dobson2018]

@@

1.	Inter-rater reliability of nurses’ TROOPS ratings; and
2.	Time taken for nurses to complete the TROOPS tool (calculated to determine the impact that integrating auditing of sedation outcomes into routine clinical practice on nurse workload).

@@

Patients undergoing procedures with nurse-administered sedation in a cardiac catheterization laboratory were eligible. 'Nurse-administered sedation' is when nurses administer a combination of sedative and analgesic medications prescribed by the proceduralist performing the procedure.[@conway2013a] A convenience sample was used.

@@

The second part consisted of the quality improvement version of the TROOPS tool, with the addition of minor adverse events from the 'airway and breathing' as well as 'gastrointestinal' domains from the research version of the tool.[@roback2018tracking] Minor events from the 'airway and breathing' domain used were 'increased or added supplemental oxygen', 'airway repositioning' and 'tactile stimulation'. Based on consultation with members of the clinical team, we elected not to include the minor items related to hypersalivation (i.e., suctioning for hypersalivation and anticholinergic for hypersalivation) because the medications used for sedation in the department (i.e., fentanyl and midazolam) are not associated with this particular adverse effect. Both minor events from the gastrointestinal domain of the research version were used. These were administration of anti-emetic and suction for emesis. The minor items from the allergy and neurological domains were not included because these adverse effects were not considered to occur frequently enough in this particular sedation context to warrant inclusion as a measure for quality improvement. Other modifications were made to the TROOPS tool through consultation with members of the clinical team in order to make it suitable for use by nurses in that particular department. Modifications included the removal of outcomes that were not feasibly assessed by nurses in the immediate post-procedure period, including permanent neurological deficit from the sentinel category of the neurological domain and pulmonary aspiration syndrome from the sentinel adverse events category of the airway and breathing domain. The TROOPS tool has not undergone validity or reliability assessment in any other clinical setting.

@@

Two nurses who performed the ‘scout’ and ‘monitor’ role during procedures performed with nurse-administered sedation completed the modified version of the TROOPS tool. The nurse performing the 'scout' role was responsible for administering direct patient care and procedural support. The nurse performing the 'monitor' role was responsible for physiological monitoring and related documentation. Ratings were performed independently as soon as possible after the completion of the procedure. Prior to commencement of the project, all nurses involved in administering sedation in the clinical department were invited to attend in-services where information about the project and the assessment tool was provided. We did not collect information on the number of nurses attending the education sessions.

@@

Data were collected from January to November, 2017. Two nurses independently completed the modified version of the TROOPS tool for 40 patients who underwent a procedure with sedation. A summary of the patient characteristics is presented in Table \@ref(tab:tab1). Most patients were undergoing a cardiac implantable electronic device implant (e.g. pacemaker or implantable cardioverter defibrillator) or electrophysiology procedure (e.g., atrial fibrillation or ventricular tachycardia ablation). A combination of midazolam and fentanyl was used for all patients.

A sedation-related adverse event was selected by at least one of the raters for 21 of the patients (52%; 95% CI = 0.38 to 0.67). Most were minor events related to the airway and breathing category of the TROOPS tool (n = 17; 42%; 95% CI = 0.29 to 0.58). The remaining were intermediate-severity events related to sedation quality and patient experience. No events for the other categories of the TROOPS tool were selected. The intra-class correlation coefficient between paired nurse ratings of the modified TROOPS tool was `r total` (95% CI = `r total_lower` to `r total_upper`). Responses from fifty-one (85%, 95% CI = 73% to 92%) nurses indicated that it took less than 1 minute to complete the modified TROOPS tool. The remaining nurses responded that it took between 1 and 3 minutes to complete. There were 20 missing responses to this item for the 40 patients.

@@

Reliability is a vital property of a measurement tool if results are intended to inform quality improvement.  There was moderate consistency between paired ratings of the modified version of the TROOPS tool used in this study. However, it should be noted that confidence intervals for the intra-class correlation were wide, spanning from poor (`r total_lower`) to very high levels of consistency (`r total_upper`). Additional research with a larger sample size would increase confidence in estimates of inter-rater reliability.

@@

We did not collect information on which nurses completed ratings so that anonymity was maintained. As such, the total number of nurses who completed ratings for patients in this study is not known. For the same reason, we can not report on the characteristics of the nurses who provided ratings, such as their experience or training.[@kottner2011guidelines] It should also be noted that, as expected, most events identified were from the minor category of the airway and breathing domain. Results should not be generalized outside of this context. The less common and more severe adverse events, such as the unintended requirement for tracheal intubation during sedation, is not likely to be inconsistently rated by nurses. Likewise, results should be considered most applicable to the context in which sedation was performed in this study. An approach that is common across many cardiac catheterization laboratories was used, with bolus doses of midazolam and fentanyl being administered by nurses.[@Conway_2014] Although there was a substantial proportion of events recorded by nurses in the sample (52%), it is possible that it over-estimates the true event rate because the analysis was performed using a convenience sample. The sample represents only a small proportion of the total number of procedures performed with sedation at the site over the data collection period. Estimating the prevalence of adverse sedation-related events was not the primary purpose of this study so we did not collect information about the total number of procedures performed with sedation over the data collection period to determine the exact proportion of completed TROOPS ratings. Finally, estimates of inter-rater reliability reported in this study should only be considered applicable to the modified version of the tool that we used, not the original research or QI versions of the TROOPS tool.

@@

Paired ratings of sedation-related adverse events by nurses in the cardiac catheterization laboratory achieved moderate consistency using a modified version of the TROOPS tool containing items from the minor category of the airway and breathing domain. Use of this tool in clinical practice seems feasible. The time required to complete the ratings was very minimal, with most ratings taking less than one minute to complete.
